Characteristic polynomials of p-adic matrices

### 2017-02-04 charpoly/charpoly.tex
ISSAC 2017 template, intro to Sec 5, 3.4 updated.

\CopyrightYear{2016}
\setcopyright{acmcopyright}
\conferenceinfo{ISSAC '16,}{July 19-22, 2016, Waterloo, ON, Canada}
\isbn{978-1-4503-4380-0/16/07}\acmPrice{\$15.00}
\doi{http://dx.doi.org/10.1145/2930889.2930898}

@@

Proposition~\ref{prop:shortcom} shows that $\com(X{-}M)$ can be encoded 
by the datum of the quadruple $(\alpha, P, Q, \chi_M)$ whose total size 
stays within $O(n^2)$ : the polynomials $\alpha$ and $\chi_M$ are 
determined by $2n$ coefficients while we need $2n^2$ coefficients to 
write down the matrices $P$ and $Q$. 
We shall see moreover in \S \ref{sec:eigenvalues} that interesting
informations can be read off on this short form $(\alpha, P, Q, 
\chi_M)$.

@@

\noindent 3.\  \:  \textbf{Eliminate}, modulo $X^{n+1}$ the coefficients $U_{i,j},$ for $j\geq i+1$ 
using the invertible pivot
$U_{i,i}=1+XL_{i,i} \mod X^{n+1}$ (with $L_{i,i} \in \Zp [X]$).

@@

\begin{theo}
Let $H \in M_n(O_K)$ be an approximate Hessenberg matrix.
Then, using Algorithm 2, one can compute $\com (X -H)$ in 
$\softO (n^3)$ operations in $O_K$ at the precision given by $H.$
\end{theo}
\begin{proof}
Firstly, the operations of the lines 2 and 3 are in $\softO (n^3)$ operations
in $O_K$ at the precision given by $H.$
Indeed, since $H$ is an approximate matrix, when we use $U_{i,i}$ as pivot:
on its columns, only $U_{i+1,i}$ can be a non zero-approximate coefficient.
As a consequence, when performing this column-pivoting, only two rows ($i$ and
$i+1$) lead to operations in $O_K [[K]]$ other than checking precision.
Hence, line 3 costs $\softO (n^2)$ for the computation of $U.$
Following line 1, the computation of $Q^{-1}$ is done by operations on rows, starting from the identity matrix.
Corresponding coefficient for the corresponding shear matrix
 is the opposite of the one defined by pivoting
and the indices for the receiving and giving rows are swapped compared to the operations on columns.
Thus, we can show that it requires no additional operation in $O_K[[X]].$ $Q^{-1}$ is just filled as
an upper-triangular matrix.
This proves that the resulting cost for lines 2 and 3 is indeed in $\softO (n^3)$ operations.

@@

Thanks to the fact that the $P$ only corresponds to the product of $n-1$ 
shear matrices, the
product on line 8 is in $\softO (n^3).$
We emphasize that no division has been done throughout the algorithm.
Line 9 is costless, and the result is then proved.
\end{proof}

@@

@@

\begin{theo}
Algorithm 3 compute 
$\com (X-M) \mod (\chi(M))$ in average complexity $\softO (n^3)$
operations in $O_K$ at initial precision.
\end{theo}
\begin{proof}
As we have already seen, completing Step 1 and 2 is in $\softO (n^3).$
Multiplying by $T$ or $S$ or their inverse corresponds
to $n$ operations on rows or columns over a matrix with coefficients
in $O_K[X]$ of degree at most $n.$
Thus, it is in $\softO (n^3).$
Step 5 and 6 are also in $\softO (n^3).$
All that is to prove is that the set of $P$ and $S$ to avoid
is of dimension at most $n-1.$
As in the proof of \ref{prop:factor_com}, we can work modulo $X-\lambda$
for $\lambda$ a root of $\chi (M)$ (in an algebraic closure)
and then apply Chinese Remainder Theorem.
The goal of the Step $2$ is to ensure the first row of $B$ contains an
invertible modulo $\chi (M).$
Since $A(\lambda)$ is of rank one, the $\mu_i$'s have to avoid an
affine hyperplane so that $row(B,1) \mod (X-\lambda)$ is a non-zero vector.
Hence for $row(B,1) \mod \chi (M)$ to contain an invertible coefficient,
a finite union of affine hyperplane is to avoid.
Similarily, the goal of Step 3 is to put an invertible coefficient (modulo
$\chi(M)$) on $C_{1,1},$ and again, only a finite union of affine
hyperplane is to avoid.
Hence, the set that the $\mu_i$'s have to avoid is a finite union
of hyperplane, and hence, is of dimension at most $n-1.$
Thus, almost any $\mu_i$ lead to a matrix $C$ passing the test
on Step 4, which concludes the proof.
\end{proof}

@@

\todo{Write a small introduction}

>>>>

\CopyrightYear{2017}
\setcopyright{acmcopyright}
\conferenceinfo{ISSAC '17,}{July 25-28, 2017, Kaiserslautern, Germany}
\isbn{***-*-****-****-0/17/07}\acmPrice{\$15.00}
\doi{http://dx.doi.org/10.1145/*******.*******}

@@

Proposition~\ref{prop:shortcom} shows that $\com(X{-}M)$ can be encoded 
by the datum of the quadruple $(\alpha, P, Q, \chi_M)$ whose total size 
stays within $O(n^2)$ : the polynomials $\alpha$ and $\chi_M$ are 
determined by $2n$ coefficients while we need $2n^2$ coefficients to 
write down the matrices $P$ and $Q$. 
We shall see moreover in Subsection \ref{ssec:eigenvalues} that interesting
informations can be read off on this short form $(\alpha, P, Q, 
\chi_M)$.

@@

\noindent 3.\  \:  \textbf{Eliminate}, modulo $X^{n+1}$ the coefficients $U_{i,j},$ for $j\geq i+1$ 
using the invertible pivot
$U_{i,i}=1+XL_{i,i} \mod X^{n+1}$ (with $L_{i,i} \in O_K[X]$).

@@

\begin{theo}
Let $H \in M_n(O_K)$ be an approximate Hessenberg matrix.
Then, using Algorithm 2, one can compute $\com (X -H)$
and $\chi_M$ in 
$\softO (n^3)$ operations in $O_K$ at the precision given by $H.$
\end{theo}
\begin{proof}
Firstly, the operations of the lines 2 and 3 are in $\softO (n^3)$ operations
in $O_K$ at the precision given by $H.$
Indeed, since $H$ is an approximate matrix, when we use $U_{i,i}$ as pivot:
on its columns, only $U_{i+1,i}$ can be a non zero-approximate coefficient.
As a consequence, when performing this column-pivoting, only two rows ($i$ and
$i+1$) lead to operations in $O_K [[K]]$ other than checking precision.
Hence, line 3 costs $\softO (n^2)$ for the computation of $U.$
Following line 1, the computation of $Q^{-1}$ is done by operations on rows, starting from the identity matrix.
Corresponding coefficient for the corresponding shear matrix
 is the opposite of the one defined by pivoting
and the indices for the receiving and giving rows are swapped compared to the operations on columns.
Thus, we can show that it requires no additional operation in $O_K[[X]].$ $Q^{-1}$ is just filled as
an upper-triangular matrix.
This proves that the resulting cost for lines 2 and 3 is indeed in $\softO (n^3)$ operations.

@@

Thanks to the fact that the $P$ only corresponds to the product of $n-1$ 
shear matrices, the
product on line 8 is in $\softO (n^3).$
We emphasize that no division has been done throughout the algorithm.
Line 9 is costless, and the result is then proved.
\end{proof}
\begin{rem}
If $M \in M_n(K)$ does not have coefficients in $O_K,$
then, by an appropriate scaling, we can still use
Algorithms 1 and 2 to get $\chi_M$
in $\softO (n^3)$ operations in $O_K$ at some shift
of the initial precision.
\end{rem}

We will see in Section \ref{sec:optjagged} that for an
entry matrix with coefficients known at flat precision,
Algorithms 1 and 2 are enough to
know the optimal jagged precision on $\chi_M.$

@@

In this Subsection, we are interested in a special
case of the computation of the comatrix of $X-M,$
when $\chi_M$ is square-free (which is the 
generic case).
We observe that it is possible to combine
the result of Proposition \ref{prop:shortcom}
with Algorithm 2 to actually compute
the comatrix of $X-M.$
It leads to a $\softO (n^3)$ algorithm
where the only divisions are for
gcd and modular inverse computations.

@@

\begin{theo}
For $M \in M_n(O_K)$ such that $discr( \chi_M) \neq 0,$
Algorithm 3 compute 
$\com (X-M) \mod (\chi(M))$ in average complexity $\softO (n^3)$
operations in $K$ at initial precision.
The only divisions are in steps 4 and 5.
\end{theo}
\begin{proof}
As we have already seen, completing Step 1 and 2 is in $\softO (n^3).$
Multiplying by $T$ or $S$ or their inverse corresponds
to $n$ operations on rows or columns over a matrix with coefficients
in $O_K[X]$ of degree at most $n.$
Thus, it is in $\softO (n^3).$
Step 5 and 6 are also in $\softO (n^3).$
All that is to prove is that the set of $P$ and $S$ to avoid
is of dimension at most $n-1.$
The idea is to work modulo $X-\lambda$
for $\lambda$ a root of $\chi (M)$ (in an algebraic closure)
and then apply Chinese Remainder Theorem.
The goal of the Step $2$ is to ensure the first row of $B$ contains an
invertible modulo $\chi (M).$
Since $A(\lambda)$ is of rank one, the $\mu_i$'s have to avoid an
affine hyperplane so that $row(B,1) \mod (X-\lambda)$ is a non-zero vector.
Hence for $row(B,1) \mod \chi (M)$ to contain an invertible coefficient,
a finite union of affine hyperplane is to avoid.
Similarily, the goal of Step 3 is to put an invertible coefficient (modulo
$\chi(M)$) on $C_{1,1},$ and again, only a finite union of affine
hyperplane is to avoid.
Hence, the set that the $\mu_i$'s have to avoid is a finite union
of hyperplane, and hence, is of dimension at most $n-1.$
Thus, almost any $\mu_i$ lead to a matrix $C$ passing the test
on Step 4.
Now, it is clear that the only divisions happening are on steps 4 and 5.
This concludes the proof.
\end{proof}

\begin{rem}
As in the previous section, it is possible
to scale $M \in M_n(K)$ so as to
get coefficients in $O_K$ and apply the previous algorithm.
\end{rem}
\begin{rem}
We refer to \cite{caruso:15a} for the handling
of the precision on gcd and modular inverse computations.
In this article, ways to tame the loss of precision
coming from divisions are explored, following
the methods of \cite{caruso-roe-vaccon:14a}.
\end{rem}

@@

In the previous Sections, \ref{sec:diffHess} and \ref{sec:diffHess},
we have proposed algorithms to obtain the comatrix of 
$X-M.$ Our motivation for these computations is to then
be able to understand what is the optimal precision on $\chi_M.$
In this section, we provide some answers to this question of
the optimal precision on $\chi_M,$ along with
numerical evidence.
We also show that it is then possible to derive 
optimal precision on eigenvalues of $M.$
